ABM simulating short-term mating behaviors

### 2022-05-30 submission/manuscript.Rmd
edits in writing

Gender differences in short-term mating behaviors are well-documented in human sexuality research. Existing studies usually conflate gender differences in mating preferences with differences in sexual behaviors, which is theoretically problematic. Using an agent-based model, we investigated the circumstances under which heterosexual and homosexual men and women’s differential preferences for short-term mating would result in gender differences in short-term mating behaviors. The model showed that when all individuals in a closed heterosexual population were considered,  men and women had the same average number of short-term mating experiences and short-term mates even when men had stronger preferences for short-term mating. Men (vs. women) had a higher average number of both experiences and mates only when heterosexual men and women in the mating pool (i.e., those with a non-zero number of short-term mating experiences) were considered. Moreover, when men had stronger preferences for short-term mating than women, gay men had a higher average number of both experiences and mates compared to both lesbian women and heterosexual men. These results suggest that theoretically speaking, even when gender differences in mating preferences existed, those in short-term mating behaviors only occurred among particular populations, or when men’s preferences for short-term mating are not constrained by those of women. Suggestions for future research in human mating psychology and behaviors were provided.

@@

Gender differences in sexual behaviors, especially short-term mating behaviors, are well-documented in human sexuality research. For example, research has found that heterosexual men, as compared to heterosexual women, have a higher average number of past short-term sexual partners [@oliver_gender_1993; @petersen_meta-analytic_2010; @rissel_heterosexual_2014], and engage in both short-term mating and extramarital sex more often [@petersen_meta-analytic_2010]. These gender differences have also been observed between gay men and lesbian women [@bailey_effects_1994; @bryant_relationship_1994; @peplau_national_1997; @peplau_sexuality_2004].

@@

Intuitive as this line of reasoning is, there are good reasons to doubt its soundness. This is because any heterosexual sexual encounter involves both a man and a woman [the frequency of encounters involving more than two persons is negligible, e.g., @herbenick_sexual_2017]. A new short-term mating experience for a man is, therefore, also a new experience for a woman, and the same logic applies to counting a new short-term mate. As a result, the psychological differences in short-term mating preferences may not result in behavioral ones in the heterosexual case [@archer_reality_2019].

As a starting point, we took gender differences in mating preferences as conceptualized by the sexual strategy theory [@buss_sexual_1993] and repeatedly supported by empirical investigations [e.g., @schmitt_universal_2003; @walter_sex_2020]. Using an agent-based model, we investigated whether men and women's differential preferences for short-term mating would result in gender differences in short-term mating behaviors, and if so, then under what conditions they would result in differences in the number of short-term mating experiences and short-term mates.

@@

As for mating standards, men are less selective in accepting someone as a potential short-term mate. For example, studies using U.S. college samples have found that the minimum percentile ranks that men found acceptable for a potential short-term sexual partner were lower than those accepted by women, both in terms of overall desirability and individual traits (e.g., social status, attractiveness) [@kenrick_evolution_1990; @regan_minimum_1998]. Another study also found that when presented with identical descriptions of potential mates, men on average rated them as more desirable than women did [@wiederman_evolution_1998].

Some evidence shows that gender differences in mating preferences also existed between gay men and lesbian women. A survey study using a community sample from the U.S. found that gay men were more interested in short-term mating than lesbian women were, and that this gender difference was comparable to that existing among heterosexual individuals [@bailey_effects_1994]. An additional indirect piece of evidence on differential standards for short-term mates comes from a recent study finding that significantly more gay men than lesbian women reported to have accepted a casual sexual offer from a same-gender person [@matsick_gender_2021]. Past studies suggest that the gender difference in the acceptance rate of casual sexual offers may originate from men and women’s differential standards for short-term mates [@conley_women_2011; @hald_receptivity_2010]. Thus, as a postulation, gay men may also have lower standards than lesbian women for short-term mates.

## Constraints on men's preferences for short-term mating
Since heterosexual sex in most cases involves one man and one woman, men's relatively high interest in short-term mating can be constrained by women's preference for long-term relationships [@archer_reality_2019; @symons_evolution_1979]. This is because men’s short-term mating preferences – with the relatively rare exception of sexual violence – can only translate into behaviors when there are women willing to have sex with them. When a new short-term mating encounter occurs, it counts both towards men’s total number of short-term mating encounters as well as towards women’s. Similarly, a new short-term mate counts towards both men and women’s total number of short-term mates. Therefore, the total number of short-term mating experiences and short-term mates must be equal between heterosexual men and women at the population level. However, it is important to note that although heterosexual men’s total number of short-term mating is equal to that of women, this does not necessarily mean that the total number of men and women who have ever had short-term mating must be equal. It is possible that there is a smaller proportion of men (vs. the proportion of women) who contribute to the total number of short-term mating. Following this line of reasoning, we would expect to observe no gender difference in short-term mating behaviors among heterosexual individuals even when there are gender differences in mating preferences.

@@

By using a spatial agent-based model, the present study investigated whether men and women’s differential preferences for short-term mating would result in gender differences in short-term mating behaviors, and if they did, under what circumstances. Preferences for short-term mating were operationally defined as the likelihood of engaging in short-term mating when encountering a mate as well as the standards for short-term mates [@buss_sexual_1993]. Short-term mating behaviors were operationally defined as the number of short-term mating experiences and the number of past short-term mates [@petersen_meta-analytic_2010].

We modeled this process among both heterosexual individuals and gay men and lesbian women to examine whether constraints set by the opposite sex’s preferences would affect short-term mating behaviors. Individuals’ sexual orientation was conceptualized in terms of behaviors only in our model. Heterosexual men had short-term mating or formed a long-term relationship only with women, while gay men only with other men, and mutatis mutandis for women.

@@

The likelihood of engaging in short-term mating and the standard for short-term mates were initialized either identically or differently for men and women, depending on the experimental conditions.

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=100mm]{figures/fig1_men_vs_women.pdf}
  \caption{\textbf{Short-term mating behaviors of men and women after 1,000 time steps in the model when gender differences existed in mating preferences.} Violin plots summarizing the outcome variables separately for heterosexual individuals and gay men and lesbian women. Plot (A) shows the average number of short-term mating experiences, and plot (B) shows the average number of short-term mates. Central points show mean values and whiskers represent standard errors, but the standard errors are small and overlap to form a single bar. Statistics were calculated using the full population of men and women in the model (\textit{N\textsubscript{men}} = \textit{N\textsubscript{women}} = 150).}
\end{figure}

@@

Gay men engaged in short-term mating behaviors more than heterosexual men did (Figure 2). The average number of short-term mating experiences was higher among gay men (*M* = `r m_exp_all.mean_1_11`, *SD* = `r m_exp_all.sd_1_11`) than among heterosexual men (*M* = `r m_exp_all.mean_0_11`, *SD* = `r m_exp_all.sd_0_11`), *t*(`r exp.all.parameter_hg`) = `r exp.all.statistic_hg`, *p* `r exp.all.p.value_hg`, Cohen's *d* = `r exp_all.d_hg`. Similarly, the average number of short-term mates was higher among gay men (*M* = `r m_partner_all.mean_1_11`, *SD* = `r m_partner_all.sd_1_11`) than among heterosexual men (*M* = `r m_partner_all.mean_0_11`, *SD* = `r m_partner_all.sd_0_11`), *t*(`r partner.all.parameter_hg`) = `r partner.all.statistic_hg`, *p* `r partner.all.p.value_hg`, Cohen's *d* = `r partner_all.d_hg`.

\begin{figure}[h]
  \centering
  \includegraphics[width=0.8\columnwidth]{figures/fig2_hetero_vs_gay_men.pdf}
  \caption{\textbf{Short-term mating behaviors of heterosexual and gay men after 1,000 time steps in the model when gender differences existed in mating preferences.} Violin plots summarizing the outcome variables. Plot (A) shows the average number of short-term mating experiences, and plot (B) shows the average number of short-term mates. Central points show mean values and whiskers represent standard errors, but the standard errors are small and overlap to form a single bar. Statistics were calculated using the full population of heterosexual and gay men in the model (\textit{N\textsubscript{hetero}} = \textit{N\textsubscript{gay}} = 150).}
\end{figure}

@@

```{r table_1, include=TRUE, echo=FALSE}
results %>%
  filter(hls %in% c("0_0_0", "0_0_1", "0_1_0", "0_1_1")) %>%
  select(m_exp_ip.mean, m_exp_ip.sd,
         f_exp_ip.mean, f_exp_ip.sd, exp_ip.d,
         m_partner_ip.mean, m_partner_ip.sd,
         f_partner_ip.mean, f_partner_ip.sd, partner_ip.d) %>%
  mutate(
    exp_ip.d = abs(exp_ip.d),
    partner_ip.d = abs(partner_ip.d)
  ) %>%
  mutate(across(where(is.numeric), inline_hook)) %>%
  mutate(
    condition =
      c("Same likelihood*Same standard",
        "Same likelihood*Different standard",
        "Different likelihood*Same standard",
        "Different likelihood*Different standard")
  ) %>%
  column_to_rownames("condition") %>%
  kable(
    digits = 2, format = "latex", booktabs = TRUE,
    caption = "\\textbf{Short-term mating behaviors of heterosexual men and women in the mating pool after 1,000 time steps in the model}",
    row.names = TRUE, col.names = c(rep(c(rep(c("M", "SD"), 2), "d"), 2))
    ) %>%
  row_spec(0, italic = TRUE, align = "c") %>%
  add_header_above(
    c(" " = 1,
      "Men" = 2, "Women" = 2, " " = 1,
      "Men" = 2, "Women" = 2, " " = 1)
    ) %>%
  add_header_above(
    c(" " = 1,
      "Average short-term mating\nexperiences (in-pool)" = 5,
      "Average short-term\nmates (in-pool)" = 5)
    ) %>%
  column_spec(1, width = "2in") %>%
  footnote(
    general = "\\\\textit{Note.} The statistics were calculated with subsamples of men and women who had had engaged in short-term mating behaviors. See Supplementary Materials for the mean and standard deviation of sample sizes.",
    general_title = "", escape = FALSE,
    threeparttable = TRUE, fixed_small_size = TRUE) %>%
  kable_styling(latex_options = "scale_down")
```

@@

Moreover, gender differences in short-term mating behaviors emerged when heterosexual men and women had different mating standards, but not when they had different short-term mating likelihood. When women had a higher standard than men did, less men than women in the population were above a potential partner’s standard and thus had a chance to have sex with them. This contributed to the unequal number of men and women in the mating pool, which led to gender differences in short-term mating behaviors among this population.

In the light of these results, previous empirical observations of gender differences in short-term mating behaviors among heterosexual individuals [e.g., @petersen_meta-analytic_2010] appear perplexing [e.g., @s_j_gurman_six_1989]. Possible explanations for these empirical results that seem illogical considering the simulation findings may have to do with features of the observation process. One possibility is sampling bias in the observations [e.g., @wiederman_evolution_1998] since surveys regarding short-term mating behaviors may tend to attract individuals who already engage in such behaviors. Another possibility is that heterosexual men and women tend to report short-term mating behaviors differently. This difference may be a result of social desirability bias [@alexander_truth_2003] or due to men and women's different estimation strategies (e.g., men tend to round up) [@brown_estimating_1999].

@@

Using agent-based modeling, the present study simulated the circumstances under which men and women’s differential preferences for short-term mating resulted in gender differences in short-term mating behaviors. We found when men (vs. women) had a stronger preferences for short-term mating, heterosexual men and women engaged in short-term mating behaviors to the same extent, while gay men engaged in more short-term mating behaviors as compared to lesbian women. Gay men engaged in more short-term mating behaviors than heterosexual men did.

>>>>

Gender differences in short-term mating behaviors are well-documented in human sexuality research. Existing studies usually conflate gender differences in mating preferences with differences in sexual behaviors, which is theoretically problematic. Using an agent-based model, we investigated the circumstances under which heterosexual and homosexual men and women’s differential preferences for short-term mating would result in gender differences in short-term mating behaviors. The model showed that when all individuals in a closed heterosexual population were considered,  men and women had the same average number of short-term mating experiences and short-term mates even when men had stronger preferences for short-term mating. Men (vs. women) had a higher average number of both experiences and mates only when heterosexual men and women who successfully participated in the mating pool (i.e., those with a non-zero number of short-term mating experiences) were considered. Moreover, when men (vs. women) had stronger preferences for short-term mating, gay men had a higher average number of both experiences and mates compared to both lesbian women and heterosexual men. These results suggest that theoretically speaking, even when gender differences in mating preferences exist, those in short-term mating behaviors only occur among particular populations, or when men’s preferences for short-term mating are not constrained by those of women. Suggestions for future research in human mating psychology and behaviors were provided.

@@

Gender differences in sexual behaviors, especially short-term mating behaviors, are well-documented in human sexuality research. For example, research has found that heterosexual men, as compared to heterosexual women, report to have a higher average number of past short-term sexual partners [@oliver_gender_1993; @petersen_meta-analytic_2010; @rissel_heterosexual_2014], and engage in both short-term mating and extramarital sex more often [@petersen_meta-analytic_2010]. These gender differences have also been observed between gay men and lesbian women [@bailey_effects_1994; @bryant_relationship_1994; @peplau_national_1997; @peplau_sexuality_2004].

@@

Intuitive as this line of reasoning is, there are good reasons to doubt its soundness. This is because any heterosexual sexual encounter involves both a man and a woman [the frequency of encounters involving more than two persons is negligible, e.g., @herbenick_sexual_2017]. A new short-term mating experience for a man is, therefore, also a new experience for a woman. Likewise, when a man has a new short-term mate, so does his partner. As a result, the psychological differences in short-term mating preferences may not result in behavioral ones in the heterosexual case [@archer_reality_2019].

As a starting point, we took gender differences in mating preferences as conceptualized by the sexual strategy theory [@buss_sexual_1993], which is repeatedly supported by empirical investigations [e.g., @schmitt_universal_2003; @walter_sex_2020]. Using an agent-based model, we investigated whether men and women's differential preferences for short-term mating would result in gender differences in short-term mating behaviors, and if so, then under what conditions they would result in differences in the number of short-term mating experiences and short-term mates.

@@

As for mating standards, men are less selective in accepting someone as a potential short-term mate. For example, studies using U.S. college samples have found that the minimum percentile ranks of the desirability of a potential short-term sexual partner accepted by men were lower than those accepted by women, in terms of both the overall desirability and the desirability of individual traits (e.g., social status, attractiveness) [@kenrick_evolution_1990; @regan_minimum_1998]. Another study also found that when presented with identical descriptions of potential mates, men on average rated them as more desirable than women did [@wiederman_evolution_1998].

Some evidence shows that gender differences in mating preferences also exist between gay men and lesbian women, suggesting that these differences are based on individuals' gender rather than their sexual orientation. A survey study using a community sample from the U.S. found that gay men were more interested in short-term mating than lesbian women were, and that this gender difference was comparable to that existing among heterosexual individuals [@bailey_effects_1994]. An additional indirect piece of evidence on differential standards for short-term mates comes from a recent study finding that significantly more gay men than lesbian women reported to have accepted a casual sexual offer from a same-gender person [@matsick_gender_2021]. Past studies suggest that the gender difference in the acceptance rate of casual sexual offers may originate from men and women’s differential standards for short-term mates [@conley_women_2011; @hald_receptivity_2010]. Thus, as a postulation, gay men may also have lower standards than lesbian women for short-term mates.

## Constraints on men's preferences for short-term mating
Since heterosexual sex in most cases involves one man and one woman, men's relatively high interest in short-term mating can be constrained by women's preference for long-term relationships [@archer_reality_2019; @symons_evolution_1979]. This is because men’s short-term mating preferences – with the exception of sexual violence – can only translate into behaviors when there are women willing to have sex with them. When a new short-term mating encounter occurs, it counts towards both men’s total number of short-term mating encounters as well as women’s, and both men and women have new short-term mates. Therefore, the total number of short-term mating experiences and short-term mates must be equal between heterosexual men and women at the population level. However, it is important to note that although heterosexual men’s total number of short-term mating is equal to that of women, this does not necessarily mean that the total number of men and women who have ever had short-term mating must be equal. It is possible that there is a smaller proportion of men (vs. the proportion of women) who contribute to the total number of short-term mating. Following this line of reasoning, we would expect to observe no gender difference in short-term mating behaviors among heterosexual individuals even when there are gender differences in mating preferences.

@@

By using a spatial agent-based model, the present study investigated whether men and women’s differential preferences for short-term mating would result in gender differences in short-term mating behaviors, and if they did, under what circumstances. Preferences for short-term mating were operationally defined as the likelihood of engaging in short-term mating upon encountering a mate and the standards for short-term mates [@buss_sexual_1993]. Short-term mating behaviors were operationally defined as the number of short-term mating experiences and the number of past short-term mates [@petersen_meta-analytic_2010].

We also modeled forming long-term relationships as a background process. A pair of individuals had a possibility to commit to a long-term relationship upon meeting each other. We modeled forming long-term relationships as a behavioral outcome and ignored the psychological processes behind it, although we recognize gender differences may also exist in long-term mating strategy [@buss_sexual_1993].

We modeled these processes among both heterosexual individuals and gay men and lesbian women to examine whether constraints set by the opposite sex’s preferences would affect short-term mating behaviors. Individuals’ sexual orientation was conceptualized in terms of behaviors only in our model. Heterosexual men had short-term mating or formed a long-term relationship only with women, while gay men only with other men, and mutatis mutandis for women.

@@

The likelihood of engaging in short-term mating and the standard for short-term mates were initialized to the values discussed previously. Whether men and women had the same initial values depended on the experimental conditions.

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=100mm]{figures/fig1_men_vs_women.pdf}
  \caption{\textbf{Short-term mating behaviors of men and women after 1,000 time steps in the model when gender differences existed in mating preferences.} Violin plots summarizing the outcome variables separately for heterosexual individuals and gay men and lesbian women. Plot (A) shows the average number of short-term mating experiences, and plot (B) shows the average number of short-term mates. Central points show mean values and whiskers represent standard errors, but the standard errors are small and overlap to form a single bar. The averages were calculated using the full population of men and women in the model (\textit{N\textsubscript{men}} = \textit{N\textsubscript{women}} = 150).}
\end{figure}

@@

Gay men engaged in short-term mating behaviors more than heterosexual men did (Figure 2). The average number of short-term mating experiences (as calculated using the full population of men) was higher among gay men (*M* = `r m_exp_all.mean_1_11`, *SD* = `r m_exp_all.sd_1_11`) than among heterosexual men (*M* = `r m_exp_all.mean_0_11`, *SD* = `r m_exp_all.sd_0_11`), *t*(`r exp.all.parameter_hg`) = `r exp.all.statistic_hg`, *p* `r exp.all.p.value_hg`, Cohen's *d* = `r exp_all.d_hg`. Similarly, the average number of short-term mates (as calculated using the full population of men) was higher among gay men (*M* = `r m_partner_all.mean_1_11`, *SD* = `r m_partner_all.sd_1_11`) than among heterosexual men (*M* = `r m_partner_all.mean_0_11`, *SD* = `r m_partner_all.sd_0_11`), *t*(`r partner.all.parameter_hg`) = `r partner.all.statistic_hg`, *p* `r partner.all.p.value_hg`, Cohen's *d* = `r partner_all.d_hg`.

\begin{figure}[h]
  \centering
  \includegraphics[width=0.8\columnwidth]{figures/fig2_hetero_vs_gay_men.pdf}
  \caption{\textbf{Short-term mating behaviors of heterosexual and gay men after 1,000 time steps in the model when gender differences existed in mating preferences.} Violin plots summarizing the outcome variables. Plot (A) shows the average number of short-term mating experiences, and plot (B) shows the average number of short-term mates. Central points show mean values and whiskers represent standard errors, but the standard errors are small and overlap to form a single bar. The averages were calculated using the full population of heterosexual and gay men in the model (\textit{N\textsubscript{hetero}} = \textit{N\textsubscript{gay}} = 150).}
\end{figure}

@@

```{r table_1, include=TRUE, echo=FALSE}
results %>%
  filter(hls %in% c("0_0_0", "0_0_1", "0_1_0", "0_1_1")) %>%
  select(m_exp_ip.mean, m_exp_ip.sd,
         f_exp_ip.mean, f_exp_ip.sd, exp_ip.d,
         m_partner_ip.mean, m_partner_ip.sd,
         f_partner_ip.mean, f_partner_ip.sd, partner_ip.d) %>%
  mutate(
    exp_ip.d = abs(exp_ip.d),
    partner_ip.d = abs(partner_ip.d)
  ) %>%
  mutate(across(where(is.numeric), inline_hook)) %>%
  mutate(
    condition =
      c("Same likelihood*Same standard",
        "Same likelihood*Different standard",
        "Different likelihood*Same standard",
        "Different likelihood*Different standard")
  ) %>%
  column_to_rownames("condition") %>%
  kable(
    digits = 2, format = "latex", booktabs = TRUE,
    caption = "\\textbf{Short-term mating behaviors of heterosexual men and women in the mating pool after 1,000 time steps in the model}",
    row.names = TRUE, col.names = c(rep(c(rep(c("M", "SD"), 2), "d"), 2))
    ) %>%
  row_spec(0, italic = TRUE, align = "c") %>%
  add_header_above(
    c(" " = 1,
      "Men" = 2, "Women" = 2, " " = 1,
      "Men" = 2, "Women" = 2, " " = 1)
    ) %>%
  add_header_above(
    c(" " = 1,
      "Average short-term mating\nexperiences (in-pool)" = 5,
      "Average short-term\nmates (in-pool)" = 5)
    ) %>%
  column_spec(1, width = "2in") %>%
  footnote(
    general = "\\\\textit{Note.} The averages were calculated with subsamples of men and women who had had engaged in short-term mating behaviors. See Supplementary Materials for the mean and standard deviation of sample sizes.",
    general_title = "", escape = FALSE,
    threeparttable = TRUE, fixed_small_size = TRUE) %>%
  kable_styling(latex_options = "scale_down")
```

@@

Moreover, gender differences in short-term mating behaviors emerged when heterosexual men and women had different mating standards, but not when they had different short-term mating likelihood. When women had a higher standard than men did, less men than women in the population were above a potential partner’s standard and thus had a chance to have sex with them. This contributed to the unequal number of men and women in the mating pool, which led to gender differences in short-term mating behaviors among this population. In comparison, even when men and women had different short-term mating likelihood, the probability of any pair of them ending up having sex is the product of their individual likelihood (since men and women made decisions independently). As this probability is the same for both parties, the gender difference in short-term mating likelihood did not translate into different behaviors.

In the light of these results, previous empirical observations of gender differences in short-term mating behaviors among heterosexual individuals [e.g., @petersen_meta-analytic_2010] appear perplexing [e.g., @s_j_gurman_six_1989]. Possible explanations for these empirical results that seem illogical considering the simulation findings may have to do with features of the observation process. One possibility is sampling bias in the observations [e.g., @wiederman_evolution_1998] since surveys regarding short-term mating behaviors may tend to attract individuals who already engage in such behaviors. Our results suggest that when this group of individuals is considered, there can be gender differences in short-term mating behaviors that do not exist in the full population. A second explanation is that heterosexual men and women's self-reported sexual behaviors are affected by social desirability bias due to gender norms. Men may overreport since having more sexual partners can show their masculinity and yield them reputation benefits, while women may underreport because this violates the chastity norm [@fisher2013gender]. A third explanation is that men and women have different estimation strategies of their sexual experiences. Men may tend to approximate and round up, while women's tendency to count instances may lead to lower estimations [@brown_estimating_1999; @mitchell_why_2019].

@@

Using agent-based modeling to simulate short-term mating behaviors, we found when men (vs. women) had a stronger preferences for short-term mating, heterosexual men and women engaged in short-term mating behaviors to the same extent, while gay men engaged in more short-term mating behaviors as compared to lesbian women. Gay men engaged in more short-term mating behaviors than heterosexual men did.
